---
title: "Introduction to Data"
author: "jasleen Hora"
date: "`r format(Sys.time(), '%F')`"
# date: "[YYYY-MM-DD]<h6>_(Last updated `r (format(Sys.time(), '%F'))`)_</h6>"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: journal
    fig_caption: yes
    css: CSS/CSS code
params:
  version: 1.0.0
---
Dataset used: High School and Beyond
We load the data using `data()` function 
Structure of data: use `str()` or `glimpse()` function

```{r setup, include=FALSE}
# packages
library(openintro)
library(dplyr)
library(ggplot2)

# knitting options
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE)

#parameters


# if data files are needed, set to TRUE
refresh_data <- FALSE
```

# Data

```{r data}
data("email50")

# structure of data
str(email50)

# or
glimpse(email50)
```

# Types of Variable
1) Numerical or Quantitative
  a) Continuous
  b) Discrete
2) Categorical or Qualitative
  a) Ordinal
  b) Nominal

# Filtering based on a factor:
Categorical data often stored as factors in R and are sometimes essential, sometimes undesirable.
Common in subgroup analysis


```{r factor }
email50_big <- email50 %>%
  filter(number == "big")

glimpse(email50_big)

table(email50_big$number)
email50_big$number <- droplevels(email50_big$number)
```

# Discretize a continuous variable:
```{r discrete}
med_num_char <- median(email50$num_char)

# creating vcategorical variable from continuos variable num_char
email50_fortified<- email50 %>%
  mutate(num_char_cat = ifelse(num_char < med_num_char, 'below median', 'at or above median'))

# count emails in each category
email50_fortified %>%
  count(num_char_cat)
```

# Another way of discretizing a continuous variable 
```{r}
email50_fortified <- email50 %>%
  mutate(number_yn = case_when(email50$number == "none" ~ "no",
                               email50$number != "none" ~ "yes"))

# Visualize number_yn
ggplot(email50_fortified, aes(x = number_yn)) + 
  geom_bar()
```

# Vizualizing Numerical Data: EDA
Scatterplot is most common to visualizerelationship between 2 numerical data. Using ggplot2. It makes modern-looking hassle free plots that take care of legends. Easy to extend code to multivariate variables. It also is easy to program and read because of its Iterative Construction.

# To check exclamation points in an email
```{r}
ggplot(email50, aes(x = num_char, y = exclaim_mess, color = factor(spam))) + 
  geom_point()
```

# Observational Studies and Experiments
1) Observationla Study:
Collect data in a way that does not directly interfere with how that data arise. Only correlation(Association?) can be inferred.( Just observed tha samples , can't contribute to the hypothesis as all factors were not considered)

2) Experimental Study:
Randomly assign subjects to various treatments. Causation(Other confounding variables that contribute to the hypothesis were also considered as we did a sampling(random assignment)) can be inferred.

# Random Sampling and Random Assignment
Random Sample: Subjects are selected randomly from the populaion and studies from the results can be generalized to that population (GENERALIZED)
Random Assignment: Allows for causal conclusion. Occurs only in experimental setings, where subjects are being assigned to various treatments.(CAUSAL) - only limited to that sample(causation)

Even if the association is formed between explanatory variable and response variable doesn't mean a causal relationship. #Simpson's paradox: The inclusion of a 3rd variable can change the relationship between explanatory and response variable.

# Sampling:
1) Simple Random Sample
2) Stratified Sample
3) Cluster Sample
4) Multistage Sample

# Principles of Experimental Design
1) Control: compare treatment of interest to a control group
2) randomize: randomly assign subjects to treatments
3) Replicate: collect a sufficiently large sample within a study, or replicate the entire study
4) Block: account for potential effect of known or suspected confounding variables. 
  Group subjects into blocks based on these variables
  Randomize within each block to treatment groups

```{r}
library(gapminder)
library(dplyr)

data("gapminder")

glimpse(gapminder) # Observational

gapminder_counts <- gapminder %>%
  count(country, continent) # can use count() or group_by()

#ucb_admission_counts %>%
  group_by(Gender) %>%
  mutate(prop = n/sum(n)) %>%
  filter(Admit == "Admitted")
  
#ucb_admission_counts %>%
  group_by(Dept, Gender) %>%
  mutate(prop = n/sum(n)) %>%
  filter(Gender == "Male" & Admit == "Admitted")

```


